Can historical data predict population responses to climate change experiments?

### 2016-12-22 precip/manuscript/precip_ms.tex
almost done with the discussion

\begin{enumerate}
\item Climate is an important driver of population ecology however there have been few tests of whether observational data correlating population performance with climate variation can be used to predict future responses of populations to experimentally imposed climate.   
\item Using longterm historical observational data from a sagebrush steppe plant community, previous research has reported a wide variety of climate effects on four dominant plant species. We tested whether these climate effects observed in observational data were useful for predicting each species response to a drought and irrigation experiment. 
\item We used rainfall shelters and automatic sprinklers to manipulate the soil moisture in sixteen new plots established at the same field site as the original historical plots. We also re-monitored the original plots as controls. After running the experiment for five years, we analyzed how the growth, survival and recruitment of each species responded to the experimental drought and irrigation treatments.   
\item In order to test whether we could predict the observed responses to the experiment we fit two types of models to the observational data collected prior to the experiment: first we fit a no climate model that included the effects of plant size and competition on plant performance in each plot in each year; next we fit a climate model that in addition to size and competition also included the effects of three seasonal soil moisture variables. Next we used these models to predict the survival and growth of each individual plant, and the total recruitment in each plot during the course of the experiment. We also used an individual based population model parameterized with the vital rate predictions to predict one step ahead changes in population size in the experimental plots. We then compared the predictions made by each of these models to the observed experimental responses and compared the performance of the no climate model to the climate model.   
\item Over the course of the experiment, average cover of the perennial grasses, \textit{Hesperostipa comata} and \textit{Pseudoroegneria spicata} declined significantly in the drought treatment. At the level of individual vital rates, experimental drought reduced the survival of \textit{Hesperostipa comata} and \textit{P. spicata}  and the growth of the grass \textit{Poa secunda}. In contrast, drought increased the growth of the shrub \textit{A. tripartita}. 
\item The climate better predicted the response to the experiment than the no climate model in six out of the twelve cases we examined. Across all species and vital rates, there was a strong positive correlation between the observed effects of the experiment and the effects predicted by the climate model fit to the observational data. At the population-level, the climate model made better predictions of the changes in species cover than the no climate model for two species: \textit{P. secunda} and \textit{P. spicata}. Moreover, in the drought treatments the climate model made better predictions for the changes in cover of all four species. 
\item \emph{Synthesis}: Observational climate data held valuable information for predicting species' responses to a modern climate change experiment in this ecosystem. Treatment responses often matched the direction of predicted responses even when the effects were not significant. We were better able to predict species' responses to the drought treatment than to the control and irrigation treatments. Suggesting that soil moisture is an important factor in predicting the population dynamics of these species but only when water is truly limiting.

@@

And yet even overcoming this challenge, predicting the effects of climate change would face yet another challenge: in many parts of the world, climate change will lead to mean annual temperatures that will be above the range of historical variation. If species performance responds non-linearly to temperature, we may be unable to predict how species respond to temperatures outside the range of observed variation \citep{doak_demographic_2010}. In addition, climate change will not just change the average climate, but is also likely to increase the variance in precipitation and lead to extreme events, which will have their own consequences independent of changes in climate averages \citep{drake_population_2005,jentsch_new_2007}. Any models based on observations drawn from the historical range of variation will therefore be extrapolating beyond both the range of observed averages and variance when used to predict the future \citep{williams_novel_2007}. Demonstrating that models fit to historical data can accurately predict the effects of experimental climate manipulations, especially manipulations that generate extreme conditions, would be strong confirmation that the climate effects they describe are not spurious and will hold even in the novel conditions of the future \cite{adler_can_2013}.

There is abundant evidence that plant performance shows high year to year variation, both at the level of individual growth, survival and reproduction and total population abundance. Precipitation is often especially important in driving interannual variation in net primary productivity among terrestrial plants \citep{knapp_variation_2001,hsu_anticipating_2014}. And the growth rates of trees are also often so tightly linked to annual precipitation that growth rings in their stems can serve as accurate records of historical climate variation thousands of years before the present \citep{yang_3500-year_2014}. Among smaller plants precipitation clearly affects the growth rates of small shrub and sub-shrub species \citep{srur_annual_2009,franklin_growth_2013}, and the germination and reproductive output of desert annuals is often tightly linked to annual precipitation \citep{venable_bet_2007}.

Despite the clear signs that climate drives net primary productivity at the ecosystem level and the variation in individual growth rate in many plants, there have been relatively few studies that clearly link observed climate variation to species performance in population models \citep{ehrlen_advancing_2016}. Fewer still have tested whether the population-level effects of climate variation shown in observational data can actually be used to predict the responses of plant populations to future climate variation. \citep{adler_can_2013} showed that population models based on observed correlations between plant population growth rates and precipitation did have some predictive power in describing species response to a short-term climate manipulation in a North American grassland. Three species showed responses to experimentally imposed drought and irrigation that were well predicted by population models fitted to historical observations.  However, another three species, showed responses to the experimental conditions that were not well predicted by historical observations. The authors suggested that limited replication in the historical data for two of these species and changing competitive conditions in the community may have led to the poor predictions.

@@

Our experiment showed that observational data on the response of plant populations to interannual climate variation can indeed help us predict the direction of species responses to experimental climate manipulations (fig \ref{fig:parPredictions}). This was true even though adding climate parameters to the demographic models only improved vital rate predictions for half of the models (fig \ref{table:overallPreds}). This should give us some hope that even when climate effects in demographic models fit to observational data are weak, they may contain useful qualitative information on the direction of climate effects in the future.

* discuss biggest successes--astreces in fig \ref{fig:parPredictions}
* discuss biggest mistake, P. secunda recruitment went in wrong direction.

Scaling up to the population level, the climate models only produced better one step a head predictions of the overall response of the species to the experiment for two species: \textit{P. spicata} and \textit{P. secunda} (table \ref{table:corPGR}). However adding climate effects produced better population-level predictions in the drought treatments for \textit{P. secunda} and \textit{H. comata}.  Both the survival and the growth of these species were positively affected by increased soil moisture (fig \ref{fig:climateGrowth}, \ref{fig:climateSurvival}), and so it makes intuitive sense that we would observe declines in these species cover in the drought treatment (fig \ref{fig:coverChange}).

*Among our species, we had the most success predicting the response of the three grass species and less success predicting the response of the shrub species \textit{A. tripartita}.    Why might ARTR be different?

*We also had more success predicting species response to the drought treatment than to the irrigation treatment. We also tended to see species respond more strongly to the drought treatment than they did to the irrigation treatment.  This tells us that in some cases, demographic models based on observational data may be of more use for prediction in extreme conditions than when conditions are close to average.  Our drought treatment likely created extremely low soil moisture compared to the historical average \ref{fig:seasonalVWC}. We hypothesize that this made water the most limiting resource for plants in these plots. In contrast, in the control and irrigated plots soil moisture may not have always been limiting during the course of the experiment.  It may make sense then that our climate models did not make more accurate predictions in these conditions than the no climate model.

@@

*Will our predictions be useful for the longer term? Because these species compete, one could argue that our predictions for any one species in this community will only be as good as the predictions we make for their competitors. For instance, while we observed little effect of drought on \textit{A. tripartita} in our experiment, it is possible that it will eventually respond positively to the drought treatment as cover of the grass species it comptetes with declines (fig \ref{fig:coverChange}). We know that grass species compete strongly with \textit{A. tripartita}.  However, we also know from previous work in this community that each species is more limited by intra-specific competition than by inter-specific crowding.  This fact ensures that the direct effects of climate change will generally be greater than the competition mediated indirect effects of climate change \cite{chu_direct_2016, kleinhesselink_indirect_2015}.

*Our results give us more confidence that historical observational data can in theory be used to detect and predict the demographic effects of climate change. This should encourage more researchers to try and use observational data to predict the future in both experimental and natural settings. Nevertheless, our success at predicting the short-term response of two species to a small-scale climate manipulation is not likely to be very reassuring to applied ecologists and resource managers wishing to make accurate medium to long-term quantitative predictions about the effects of climate change on the species they manage.  Clearly more work is needed to distinguish which predictions are worth having confidence in and which predictions we should have less confidence in. Towards that goal, perhaps the best way to build confidence in ecological predictions is conduct more tests like this one.

@@

\centerline{\Large \textbf{Supporting Information }}
\centerline{Adler et al., ``Weak interspecific interactions''}

>>>>

Climate is an important driver of population ecology however there have been few tests of whether observational data correlating population performance with climate variation can be used to predict how populations respond to climate change experiments. We tested whether longterm observational data relating the effects of climate variation to the survival, growth and recruitment of four common plants in a sagebrush steppe could be produce accurate predictions for how each species would respond to a drought and irrigation experiment. 
We used rainout shelters and automatic sprinklers to manipulate the soil moisture in sixteen new plots established at the same field site as the longterm observational data. After running the experiment for five years, we analyzed how the growth, survival and recruitment of each species responded to the experimental drought and irrigation treatments. We made predictions using two types of models fit to the observational data collected prior to the experiment: a no climate model that only included the effects of plant size and competition on plant performance in each plot in each year; and a climate model that also included the effects of three seasonal soil moisture variables. We then used these models to predict the performance of each species during the course of the experiment. We also generated one step a head predictions of population size in each experimental plot using an individual based population model. We compared predictions made by the the no climate and climate models to the observed experimental responses. 
Over the course of the experiment, average cover of the perennial grasses, \textit{Hesperostipa comata} and \textit{Pseudoroegneria spicata} declined significantly in the drought treatment. At the level of individual vital rates, experimental drought reduced the survival of \textit{Hesperostipa comata} and \textit{P. spicata}  and the growth of the grass \textit{Poa secunda}. In contrast, drought increased the growth of the shrub \textit{A. tripartita}. The climate model better predicted the response to the experiment than the no climate model in six out of twelve cases. Across all species and vital rates, there was a strong positive correlation between the observed effects of the experiment and the effects predicted by the climate model fit to the observational data. At the population-level, the climate model made better predictions of the changes in species cover than the no climate model for \textit{P. secunda} and \textit{P. spicata}. 
Observational climate data held valuable information for predicting species' responses to a climate change experiment in this ecosystem. Treatment responses often matched the direction of predicted responses even when the effects were not significant. We were better able to predict species' responses to the drought treatment than to the control and irrigation treatments. This suggests that soil moisture is an important factor in predicting the population dynamics of these species but only when water is truly limiting.

@@

And yet even overcoming this challenge, predicting the effects of climate change would face yet another challenge: in many parts of the world, climate change will lead to mean annual temperatures that will be above the range of historical variation. If species performance responds non-linearly to temperature, we may be unable to predict how species respond to temperatures outside the range of observed variation \citep{doak_demographic_2010}. In addition, climate change will not just change the average climate, but is also likely to increase the variance in precipitation and lead to extreme events, which will have their own consequences independent of changes in climate means \citep{gherardi_enhanced_2015,jentsch_new_2007}. Any models based on observations drawn from the historical range of variation will therefore be extrapolating beyond both the range of observed averages and variance when used to predict the future \citep{williams_novel_2007}. Demonstrating that models fit to historical data can accurately predict the effects of experimental climate manipulations, especially manipulations that generate extreme conditions, would be strong confirmation that the climate effects they describe are not spurious and will hold even in the novel conditions of the future \cite{adler_can_2013}.

There is abundant evidence that plant performance shows high year to year variation, both at the level of individual growth, survival and reproduction and total population abundance. Precipitation is often especially important in driving interannual variation in net primary productivity among terrestrial plants \citep{knapp_variation_2001,hsu_anticipating_2014}, the annual growth rings in trees \citep{yang_3500-year_2014} and smaller plants \citep{srur_annual_2009,franklin_growth_2013}, as well as the germination and reproductive output of many annuals \citep{venable_bet_2007}.

Despite the clear signs that precipitation drives net primary productivity at the ecosystem level and annual variation in individual growth rate in many plants, there have been relatively few studies that clearly link observed climate variation to species performance in population models \citep{ehrlen_advancing_2016}. Fewer still have tested whether the population-level effects of climate variation shown in observational data can actually be used to predict the responses of plant populations to future climate variation. \citep{adler_can_2013} showed that population models based on observed correlations between plant population growth rates and precipitation did have some predictive power in describing species response to a short-term climate manipulation in a North American grassland. Three species showed responses to experimentally imposed drought and irrigation that were well predicted by population models fitted to historical observations.  However, another three species, showed responses to the experimental conditions that were not well predicted by historical observations. The authors suggested that limited replication in the historical data for two of these species and changing competitive conditions in the community may have led to the poor predictions.

@@

Our experiment showed that observational data on the response of plant populations to interannual climate variation can indeed help us predict the direction of species responses to experimental climate manipulations (fig \ref{fig:parPredictions}). This was true even though adding climate parameters to the demographic models only improved vital rate predictions for half of the models (fig \ref{table:overallPreds}). This should give us some hope that even when climate effects in demographic models fit to observational data are weak or not significant, they may contain useful qualitative information on the direction of climate effects in the future.

*Although a number of previous studies of the plants in this system have looked the effects of climate on demography and survival, this is the first study to demonstrate effects of climate experimentally. The significant decline in \textit{H. comata} cover we report in the experiment (fig \ref{fig:coverChange}) is consistent with the population-level sensitivities to precipitation estimated by previous studies (\citep{chu_direct_2016},\citep{dalgleish_climate_2010}, \citep{adler_forecasting_2012}). Each of these studies report \textit{H. comata} having the strongest positive population response to increased precipitation among all the species. The response of this species' vital rates to drought in the experiment also matches this result (fig \ref{fig:growthTreat}, \ref{fig:survivalTreat}). On the other hand if we had only conducted an irrigation experiment our results may not have shown this consistency with previous work as \textit{H. comata} showed no positive response to irrigation. Previous work has also generally reported postive population responses to precipitation for the other grasses, \textit{P. secunda} and \textit{P. spicata} (\citep{adler_forecasting_2012}, \citep{chu_direct_2016}, \citep{dalgleish_climate_2010}). These results are consistent with significant decrease we saw in \textit{P. spicata} cover and survival and \textit{P. secunda} growth and survival in the drought plots in our experiment (figs \ref{fig:coverChange}, \ref{fig:growthTreat}, \ref{fig:survivalTreat}). As for \textit{H. comata} the magnitudes of the negative drought effects were greater than the magnitudes of the positive irrigation effects on these species.

In contrast to the effects on the grasses the effects on \textit{A. tripartita} are more complicated. The effects of drought and irrigation on \textit{A. tripartita} growth were highly dependent on plant size (figs \ref{fig:growthTreat}): drought had a positive effect except possibly for the largest plants (fig \ref{fig:parPredARTRGrowth}). This result matches the negative effects of precipitation on this species estimated by \citep{adler_forecasting_2012} and \citep{chu_direct_2016}. These studies also report relative strong indirect effects of precipitation on \textit{A. tripartita} mediated by its competition with grasses. It is possible that some of the positive effect on \textit{A. tripartita} growth we see in the drought plots is the result of reduced grass competition these plots (fig \ref{fig:coverPred}; \citep{chu_direct_2016}). Our growth model should take into account any changes in grass abundance that could be driving a positive response from \textit{A. tripartita}. So this leaves us with the question of why this species would show a positive direct response to drought? Although there is some evidence that saturated soils can be detrimental to sagebrush growth \citep{} there was no evidence in our data that ambient moisture would have been so wet as to be limiting to sagebrush during this experiment (fig \ref{fig:dailyVWC}). Alternatively our drought treatments could be reducing snow cover in the winter in spring and allowing for earlier growth of sagebrush.  In a high elevation sagebrush meadow earlier snowmelt was correlated with increased sagebrush performance \citep{harte_}.

One trend that appears to stand out in these results is the relatively asymmetrical effects of drought and irrigation. Although our irrigation plots raised soil moisture, we saw little positive effect of added moisture on the same species that were reduced by drought (figs \ref{fig:growthTreat}, \ref{fig:survivalTreat}). We expected that in this water limited system, increasing soil moisture would lead to a stronger increase in plant performance. But perhaps we should not have been surprised by this result. It is possible that adding water to this system does little to increase grass growth and survival for a number of reasons.  First, these species are presumably able to escape drought, by growing early in the year, or avoid drought stress through high water use efficiency \citep{bazzaz, franks}. Indeed our soil moisture data generally show a pulse of soil moisture in the spring when many grasses are actively growing (fig \ref{fig:dailyVWC}). During this window of active growth soil moisture in sagebrush steppe may be quite high and perhaps other resources become limiting \citep{caldwell} \citep{ino}. Moreover, our treatments had relatively smaller effects on soil moisture during this period \ref{table:soil_moisture_model}. If the species in our system are well adapted to the ambient water limitation, it makes some sense that they would show relatively small responses to increased moisture.

A possibly related result was recently observed by \citep{gherardi_} in a warm desert grassland: the effect of increased soil moisture on grass productivity in that system showed a concave down pattern of response to increased soil moisture.  Increasing soil moisture had weaker effects on grass productivity than decreasing soil moisture in equal proportion. This pattern could explain why we repeatedly saw large decreases in either the cover or performance of the three grasses in drought plots but did not see any statistically significant increase in performance in the irrigated plots.

Reflecting on the possible mechanisms by which soil moisture could have affected these species sheds some light on our sucesses and failures of prediction.  The drought effects we observed on the three grasses were often stronger than the effects we predicted, while the irrigation responses observed were often weaker than predicted (figs \ref{fig:parPredHECOGrowth}, \ref{fig:parPredHECOSurvival}, \ref{fig:parPredPOSEGrowth}, \ref{fig:parPredPOSESurvival}, \ref{fig:parPredPSSPGrowth}, \ref{fig:parPredPSSPSurvival}). If we had fit our growth and survival models with a non-linear function for soil moisture, perhaps informed by more mechanistic understanding of water limitation in this system, they may have made more accurate predictions of the drought and irrigation effects \citep{ehrlen_advancing_2016}.

Out of all the climate effects we predicted and observed, we only observed four cases where vital rate predictions and observations were both significantly different from zero (fig  \ref{fig:parPredictions}). In three of these cases, we successfully predicted the direction of the treatment effect. However, in one cases we predicted a treatment effect that was significantly

Using the IBM, we also demonstrated that the climate model predictions of the vital rates could be used to generate population-level predictions that were also better than the no climate models in some cases: the climate model improved one step ahead predictions of cover changes of \textit{P. spicata} and \textit{P. secunda} (table \ref{table:corPGR}). In the drought treatment our one step ahead cover predictions for \textit{H. comata} and \textit{P. secunda} were also better than the no climate model. In this analysis, we either used the climate model, or the no climate model to generate predictions for the IBM. In theory we could have refined our predictions by  climate model to predict all three vital rates in the IBM. Improvements of our cover predictions may have been possible if we only included climate models in the IBM when those models improved vital rate predictions--for instance, the \textit{A tripartita} IBM would only include the climate model for growth but not for survival and recruitment.

@@

*Will our predictions be useful for the predicting the effects of climate change in the future? Because these species compete, one could argue that our predictions for any one species in this community will only be as good as the predictions we make for their competitors \citep{tylianakis}. For instance, while we observed weak treatment effects on \textit{A. tripartita} in our experiment (fig \ref{fig:coverChange}) could drought ultimately lead to a decline in the cover of its competitors which will then allow it to increase in abundance in the future?  While this is possible for some species, research at our site and several other grasslands suggest that species are more limited by intra-specific competition than inter-specific competition \cite{chu_direct_2016,adler_weak_2016}. In theory this fact should mean that direct effects rather than indirect effects dominate species responses to climate change (\citep{levine_competitors_2010,kleinhesselink_indirect_2015,adler_forecasting_2012}).

*Our results give us more confidence that historical observational data can be used to detect and predict the demographic effects of climate change. This should encourage more researchers to try and use observational data to predict the future in both experimental and natural settings \citep{houlahan_priority_2016}. Making quantitative predictions, however humble they are, and testing them is arguably a key to scientific progress in all fields.  Nevertheless, our success at predicting the short-term response of two species to a small-scale climate manipulation is not likely to be very reassuring to applied ecologists and resource managers wishing to make accurate predictions about the effects of climate change in large complex systems.  Clearly more work is needed to distinguish which predictions are worth having confidence in and which predictions we should have less confidence in. Towards that goal, perhaps the best way forward is to conduct more tests like this one.

@@

\centerline{\Large \textbf{Supporting Information }}
\centerline{Kleinhesselink et al., ``Predicting climate response''}
